---
title: "STA522 Project 1"
author: "Holly Cui"
date: "2023-10-20"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, include=FALSE}
library(tidyverse)
library(tidyr)
library(dplyr)
library(ggplot2)
library(readxl)
library(survey)
```

```{r data}
county <- read_excel("county.xlsx", col_names = TRUE)
```

## Study Design

Stratified sampling by 50 states with optimal allocation. 

  - index: row number
  - strata: 50 state* (remove territories + DC as certainty PSU)
  - N: 3246 --> 3144 (after removal)
  - n: 342 --> need explanation*
  - Nh: # of counties in each state
  - nh: # of sampled counties in each state
  
  
## Data Cleaning

```{r}
# remove US territories
US_territories <- c("American Samoa", "Guam", "Northern Mariana Islands", 
                    "Puerto Rico", "U.S. Minor Outlying Islands", 
                    "Virgin Islands (U.S.)")

clean_county <- county %>%
  filter(!state %in% US_territories) %>%
  group_by(state) %>%
  mutate(Nh = n(), Sh = sd(pop)) %>%
  ungroup() %>%
  mutate(id = row_number())
```
  
  
## Sample Size

### Method 1 - Standard

A good maximum sample size is usually around 10% of the population. Thus, we define our sample size as $n = 0.1N = 314$. 

### Method 2 - Fancy but not applicable:

Sample size for known population: 
$$n = \frac{\frac{Z^2 \times p(1-p)}{e^2}}{1+\frac{Z^2 \times p(1-p)}{e^2N}}$$
where: 

  - $Z$: Z-score for the 95% confidence intervals
  - $p$: Standard deviation of how much individual sample data points deviate from the average population
  - $e$: Designated margin of error 
  - $N$: Population size
  
```{r}
# define parameters
Z = 1.96 # for 95% CI
p = 0.5 # default value for expected variance
e = 0.05 # usually around 2% to 5% based on accuracy
N = nrow(clean_county)

# calculate sample size n
n = round((Z^2 * p*(1-p) / e^2) / (1 + (Z^2 * p*(1-p) / (e^2 * N)))) # 342
```


## Optimal Allocation

To calculate the optimal allocation of n=343, we need to adopt the expression as follows:
$$n_h = n\times\frac{\frac{N_hS_h}{N}}{\frac{\Sigma^{H}_{h=1}N_hS_h}{N}}$$
```{r}
# state-level strata statistics (without DC - certainty PSU)
clean_state <- county %>%
  filter(!state %in% US_territories) %>%
  filter(state != "District of Columbia") %>%
  group_by(state) %>%
  summarise(Nh = n(), Sh = sd(pop)) %>%
  ungroup()
```

```{r}
# calculate denominator for optimal allocation
n = 314
denominator = sum(clean_state$Nh * clean_state$Sh)

# summarize nh for each state strata
state_strata <- clean_state %>%
  mutate(nh_round = round((n-1)*Nh*Sh / denominator)) %>%
  mutate(nh = ifelse(nh_round == 0, nh_round+1, nh_round))
         #nh_round = round(nh_true), 
         #diff = nh_round - nh_true) 

# check rounding issue (which doesn't guarantee final n = 341)
# state_strata %>%
#   arrange(desc(diff)) %>%
#   head(sum(state_strata$nh_round) - (n-1))

# fix rounding & get final sampling schema
state_nh <- state_strata %>%
  # mutate(nh_round = ifelse(state == "Tennessee", 6, nh_round), 
  #        nh_round = ifelse(state == "Pennsylvania", 9, nh_round), 
  #        nh_round = ifelse(state == "Illinois", 27, nh_round)) %>%
  select(state, Nh, Sh, nh)
```

## Weights 

```{r}
# excluding D.C. for now, its weight is 1
weights <- state_nh$nh/state_nh$Nh
```


## Sampling

```{r}
set.seed(123)

sample_county = c()
for (i in 1:nrow(state_nh)) {
  sample_by_state = sample(clean_county$id[clean_county$state == state_nh$state[i]], 
                           state_nh$nh[i])
  sample_county = c(sample_county, sample_by_state)
}

# add in DC (certainty PSU)
final_sample = c(sample_county, 322)
```

```{r}
sample = clean_county %>%
  filter(id %in% final_sample)
sample

#write.csv(sample, file = "/Users/hollycui/Desktop/STA\ 522/Project1/sample.csv", row.names = FALSE)
```

```{r}
sample <- sample %>%
  mutate(st_cty = paste(state, county, sep="")) %>%
  arrange(county)
```


# Getting Hispanic Data 

## 2020

```{r}
# estimates 7/1/2020
hispanic_data_2020 <- read.csv('demo_2010_2020.csv') %>% 
  filter(YEAR == 13, AGEGRP == 0) %>%
  select(STNAME, CTYNAME, H_MALE, H_FEMALE) 
  
head(hispanic_data_2020)
```
```{r}
# remove county from names
# weird ch in new mexico county name, not in sample
hispanic_data_2020 <- hispanic_data_2020[-c(1804),]
hispanic_data_2020$CTYNAME <- sub('County', '', hispanic_data_2020$CTYNAME)
```

```{r}
# the census data does not have Northwest Hills Planning Region
sample[!(sample$county %in% hispanic_sample_2020$CTYNAME),]
```

```{r}
hispanic_data_2020 %>%
  filter(STNAME == 'Connecticut')
```

```{r}
# keep counties in the sample
hispanic_sample_2020 <- hispanic_data_2020 %>%
  mutate(CTYNAME = trimws(CTYNAME, which = "right"),
         n_hispanic = as.numeric(H_MALE) + as.numeric(H_FEMALE),
         st_cty = paste(STNAME, CTYNAME, sep="")) %>%
  filter(st_cty %in% sample$st_cty | CTYNAME %in% c('Hawaii', 'San Francisco', 
                                                    'Wrangell City and Borough', 
                                                    'Emporia city', 'Salem city'
                                                    , 'Radford city', 'De Witt')) %>%
  arrange(CTYNAME)
```


```{r}
# removing ct county, adding data to sample df
sample <- sample[-c(204), ]
sample$hispanic_pop_2020 <- hispanic_sample_2020$n_hispanic
```


## 2010


```{r}
# estimates 7/1/2010
hispanic_data_2010 <- read.csv('demo_2010_2020.csv') %>% 
  filter(YEAR == 3, AGEGRP == 0) %>%
  select(STNAME, CTYNAME, H_MALE, H_FEMALE) 
  
head(hispanic_data_2010)
```

```{r}
# remove county from names
# weird ch in new mexico county name, not in sample
hispanic_data_2010 <- hispanic_data_2010[-c(1804),]
hispanic_data_2010$CTYNAME <- sub('County', '', hispanic_data_2010$CTYNAME)
```


```{r}
# keep counties in the sample
hispanic_sample_2010 <- hispanic_data_2010 %>%
  mutate(CTYNAME = trimws(CTYNAME, which = "right"),
         n_hispanic = as.numeric(H_MALE) + as.numeric(H_FEMALE),
         st_cty = paste(STNAME, CTYNAME, sep="")) %>%
  filter(st_cty %in% sample$st_cty | CTYNAME %in% c('Hawaii', 'San Francisco', 
                                                    'Wrangell City and Borough', 
                                                    'Emporia city', 'Salem city'
                                                    , 'Radford city', 'De Witt')) %>%
  arrange(CTYNAME)
```


```{r}
sample$hispanic_pop_2010 <- hispanic_sample_2010$n_hispanic
```
