---
title: "Project 1"
author: "Elyse McFalls and Holly Cui"
date: "2023-10-20"
output: pdf_document
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, include=FALSE}
library(tidyverse)
library(tidyr)
library(dplyr)
library(ggplot2)
library(readxl)
library(writexl)
library(survey)
```

# Data and Sampling Frame

```{r data}
county <- read_excel("county.xlsx", col_names = TRUE)
```

This project seeks to understand facets of the United States using county-level data. Specifically, we are interested in examining the characteristics of counties in the U.S., such as population density, demographic information, and political preferences. Data on the states and counties in the United States were collected from the Wikipedia page [List of states and territories of the United States](https://en.wikipedia.org/wiki/List_of_United_States_counties_and_county_equivalents) which contains the 2020 census estimates for each county's population and area. More detailed data for each county was extracted from the Unites States Census 
Bureau and MIT's Election Data and Science Labs (MEDSL). The data from [the Census](https://www.census.gov/programs-surveys/popest/technical-documentation/research/evaluation-estimates/2020-evaluation-estimates/2010s-county-detail.html) contained demographic estimates for each county from 2010 to 2020. [MEDSL](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/VOQCHQ) on the other hand, entailed data on political participation by county from 2000 to 2020.

To perform the survey, all counties from the 50 U.S. states and the District of Columbia were included in the sampling frame. However, counties from U.S. territories and remote islands were not incorporated due to the fact that these regions are often located far from the continental United States, making data collection logistically challenging and costly from a realistic point of view. Even if we are presented with the data estimates of all counties, U.S. territories and remote islands tend to have smaller populations compared to states, which can result in less statistical significance when conducting a national survey. Including these areas may not significantly impact the overall survey results. Finally, to align with one of our objectives on the topic of political preferences, we are primarily interested in regions that are enfranchised or enabled representation in the presidential votes. Since residents of these regions, such as Puerto Rico, Guam, or American Samoa, do not have full voting representation in the general elections, we decide to not consider them as part of the sampling frame for our study. 

```{r}
# Remove US territories & remote islands
US_territories <- c("American Samoa", "Guam", "Northern Mariana Islands", 
                    "Puerto Rico", "U.S. Minor Outlying Islands", 
                    "Virgin Islands (U.S.)")

clean_county <- county %>%
  filter(!state %in% US_territories) %>%
  group_by(state) %>%
  mutate(Nh = n(), Sh = sd(pop)) %>%
  ungroup() %>%
  mutate(id = row_number())
```


# Sampling Procedure

## Sampling Design

In deciding on the sampling design, we wanted to choose a framework that best fit the goals of this project. This framework would ideally give us a representative sample of the counties in the U.S. that bears in mind the variedness of each state. Therefore, we decided to conduct a simple stratified sampling with each of the 50 states serving as individual stratum. To determine the sample size in each stratum, we adopted optimal allocation to minimize the variance of estimates. In terms of the District of Columbia which operates simultaneously as a city, a county and a state, we treated it as a singleton stratum where the probability of DC being sampled equals to 1. This can be further analyzed as a [certainty sample](https://statehood.dc.gov/page/faq#:~:text=The%20Constitution%20dictates%20that%20the,4%20million%20people%20and%20growing.) with detailed treatments being explained in the estimation section. 

By taking a stratified sample, we are ensuring that every state in the U.S. is being accounted for. Stratified samples also provide estimates with lower standard errors when there is more between-variance than within-variance. We expect this to be the case with counties in each state. For instance, counties in Virginia are expected to exhibit more similarities among themselves than when compared to counties in Wyoming. However, we also believed that the variance of our variables of interest within each state may differ by state. Consider a state like North Dakota where most of the counties are rural compared to a state like North Carolina that has a mixture of rural and urban areas. We'd anticipate that [county-level population in North Dakota would be more uniform and the same statistics in North Carolina would be more varied](https://mtgis-portal.geo.census.gov/arcgis/apps/storymaps/collections/189aa1dbd64c4c81b3b4a2b71124f6c6?item=1). Therefore, instead of sampling proportional to size, we opted for an optimal allocation framework (Eq. 1). This way, we would sample more from states that have higher variances for our variables of interest. Specifically, we decided to focus on the population variable of each county to optimally allocate the samples of each stratum since we expect population values to be related to demographics. Therefore, this framework will subsequently reduce the variance in our population estimates and likely the variance in our demographic related estimates.

$$\text{Eq. 1}: n_h = n \times\frac {N_hS_h/N} {\sum^H_{h=1}N_hS_h/N}$$
where: 

  - $n_h$ is the number of sample in state $h$
  - $N_h$ is the number of counties in state $h$
  - $N$ is the total number of counties in the sampling frame
  - $n$ is the sample size
  - $H$ is the total number of states
  - $S_h$ is the variance of the county populations in state $h$

Furthermore, employing a simple random sampling approach following stratification enables us to obtain a representative sample from each state. While we contemplated adopting a sample proportional to size, we ultimately decided against it as we deemed it unnecessary to give higher priority to more densely populated areas. 

Additionally, we made the deliberate choice to include the District of Columbia (D.C.) in our sample with 100% certainty. This decision stems from D.C.'s unique status as the nation's capital, with characteristics akin to both a state and a county. If we were to treat D.C. as a state with a single county, our optimal allocation formula would assign zero samples to it, given the absence of variance in that context. Therefore, to ensure the inclusion of D.C. in our sample, we intentionally retained it as a separate certainty PSU in our sampling strategy.

```{r}
# State-level strata statistics (without DC - certainty PSU)
clean_state <- county %>%
  filter(!state %in% US_territories) %>%
  filter(state != "District of Columbia") %>%
  group_by(state) %>%
  summarise(Nh = n(), Sh = sd(pop)) %>%
  ungroup()
```


## Sample Sizes and Weights

**Sample Size (n)**: Initially, we selected a sample size of 314, equivalent to 10% of our population size (which is usually a good maximum sample size). However, when applying the optimal allocation formula, it produced instances where certain states had no counties included in the sample. To ensure representation from every state, we rounded any calculated values of $nh$ less than 0.5 up to 1. As a result, our final sample size was adjusted to 317.

The breakdown of the number of counties sampled by state using the optimal 
allocation formula is as follows:

State         | nh           |State         | nh        | State      | nh                     
------------- | -------------| -------------| ----------|----------- | -------
Alabama|3                     | Louisiana|  3             |Ohio| 9
Alaska| 1                     | Maine|  1                 |Oklahoma| 4
Arizona| 8                    | Maryland|  3              |Oregon| 3
Arkansas|2                    | Massachusetts| 4          |Pennsylvania| 9
California|39                | Michigan| 10               |Rhode Island| 1
Colorado| 5                   | Minnesota| 6              |South Carolina|  3     
Connecticut|  1               | Mississippi|  2           |South Dakota | 1                
Delaware| 1                   | Missouri| 7               |Tennessee|6
Florida| 16                    | Montana|  1               |Texas |  48                
Georgia| 11                    | Nebraska| 3               |Utah|3
Hawaii| 1                     | Nevada |  4               |Vermont|1
Idaho| 2                      | New Hampshire|1           |Virginia | 8    
Illinois|  25                  | New Jersey|  3            |Washington|7
Indiana|   5                  | New Mexico| 2             |West Virginia|1
Iowa|  3                      | New York|16                |Wisconsin|5
Kansas|  4                    | North Carolina| 8         |Wyoming|1
Kentucky| 4                | North Dakota|  1          |District of Columbia*| 1

*certainty primary sampling unit

```{r}
# Calculate denominator for optimal allocation
n = 314
denominator = sum(clean_state$Nh * clean_state$Sh)

# Summarize nh for each state stratum
state_strata <- clean_state %>%
  mutate(nh_round = round((n-1)*Nh*Sh / denominator)) %>%
  # Check rounding issue
  mutate(nh = ifelse(nh_round == 0, nh_round+1, nh_round))

# Fix rounding & get final sampling schema
state_nh <- state_strata %>%
  select(state, Nh, Sh, nh)
```

```{r}
# Sampling
set.seed(123)
sample_county = c()
for (i in 1:nrow(state_nh)) {
  sample_by_state = sample(clean_county$id[clean_county$state == state_nh$state[i]], 
                           state_nh$nh[i])
  sample_county = c(sample_county, sample_by_state)
}

# Add back DC (certainty PSU)
final_sample = c(sample_county, 322)

# Final sample
sample = clean_county %>%
  filter(id %in% final_sample)
```

**Weights**: The weight assigned to each county is determined by dividing the total number of counties within its corresponding state by the number of counties actually sampled. For example, in the case of Alabama, which has a total of 67 counties and of which 3 were included in the sample, the weight assigned to Alabama's sampled county is calculated as 67 / 3, resulting in a weight of approximately 22.33. In contrast, the District of Columbia (D.C.) carries a weight of 1 since it serves as a primary sampling unit with 100% certainty.

```{r}
# Excluding D.C. for now, its weight is 1
sample_final <- sample %>%
  left_join(state_nh, by = c("state", "Nh", "Sh")) %>%
  select(-id) %>%
  mutate(weights = Nh/nh,
         pop_density = pop/area,
         # Format city's name 
         county = ifelse(grepl("city", county, ignore.case = TRUE),
                         sub(",.*", "", county),
                         county))
```


# County-Level Data Collection

Building upon our sample of 317 counties, our next step is to acquire more detailed, county-level data. This data is essential for addressing the specific questions at hand, particularly those related to county population density, demographic compositions (such as the Hispanic or Latino population), and political affiliations. Therefore, the upcoming data collection efforts are carefully aligned with our objectives and will draw from reliable, professional sources. 

## Political Preferences

[County Presidential Election Returns 2000-2020](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/VOQCHQ) is an open-access dataset managed by MIT's Election Data and Science Labs (MEDSL). This dataset provides comprehensive records of county-level presidential election statistics spanning from the year 2000 to 2020. Each entry in the dataset corresponds to a specific county and election year, featuring the candidates of various political parties (Democrat, Republican, Green, Libertarian, and Other), along with their vote counts and the total number of votes cast in that county. Given our specific interest in the 2020 election, we have filtered out data from other years to focus the 2020 election cycle. To align this dataset with our sample dataset, we treat all parties other than Democrat or Republican as third party, and group by county and party to retrieve the number of votes for each party. Additionally, we have included the total number of votes cast in each county, enabling us to further analyze and draw insights from the 2020 election data.

Before proceeding with the integration of the retrieved values into our sample dataset, it's essential to acknowledge the considerable variability in county names across datasets. To illustrate, one of our sampled counties, *Northwest Hills Planning Region* in Connecticut, has undergone a name change from *Litchfield*. However, a majority of professional sources online still maintain the previous naming convention. As such, relying solely on a brute-force matching method may not yield accurate results. After thorough online research, we have identified a conventional solution: the use of 5-digit [Federal Information Processing Series (FIPS)](https://www.census.gov/library/reference/code-lists/ansi.html) codes to standardize county names. Since our data source also includes FIPS information, our strategy involves initially establishing a match between our sampled county names and a [FIPS dataset](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/OSLU4G). Subsequently, we will perform the data integration, linking the party data based on the FIPS codes. This approach significantly enhances our efficiency compared to the manual verification of county names.


```{r}
# Use FIPS for matching
fips <- read.csv("countyfipstool20190120.csv")

most_match <- sample_final %>%
  left_join(fips, by = join_by("county" == "cname", "state" == "sname")) %>%
  select(-c(sab, sid, sfips, saint, cfips))
```

```{r}
# Read-in fixed sample with complete FIPS
sample_match <- read_excel("sample_match.xlsx", col_names = TRUE)

# Read-in vote dataset
vote <- read.csv("countypres_2000-2020.csv")
```

```{r}
# Create vote dataframe by party and county
third_party <- c("OTHER", "GREEN", "LIBERTARIAN")

vote_by_party <- vote %>%
  filter(year == 2020) %>%
  select(state, state_po, county_name, county_fips, 
         candidate, party, candidatevotes, totalvotes) %>%
  mutate(party_group = case_when(
    party %in% third_party ~ "THIRD",
    .default = party
  )) %>%
  select(-c(party, candidate)) %>%
  group_by(state, county_name, party_group) %>%
  summarise(votes = sum(candidatevotes), .groups = "drop")

# Clean original data
vote_select <- vote %>%
  filter(year == 2020) %>%
  mutate(party_group = case_when(
    party %in% third_party ~ "THIRD",
    .default = party)) %>%
  select(state, state_po, county_name, county_fips, party_group, totalvotes) %>%
  distinct() # remove duplicate rows

# Join back for final vote data
clean_vote <- vote_by_party %>%
  left_join(vote_select, by = join_by(state, county_name, party_group)) %>%
  select(state, state_po, county_name, county_fips, party_group, votes, totalvotes) %>%
  pivot_wider(names_from = party_group, values_from = votes) %>%
  mutate(county_fips = ifelse(state_po == "DC", 11001, county_fips))
```

Following our final attempt to perform the match-back, we encountered an issue with *Wrangell* in Alaska, which was not documented in the vote data. To address this discrepancy, we turned to the [official data source](https://www.elections.alaska.gov/results/20GENR/data/sovc/HD36p.pdf), specifically the table available on page 2 of Alaska's government website, for a manual imputation process. This step was necessary to ensure the completeness and accuracy of our dataset, particularly for the county of Wrangell in Alaska.

```{r}
# Match vote with sample data
sample_w_vote <- sample_match %>%
  left_join(clean_vote, by = join_by("fips" == "county_fips")) %>%
  # Manual fix Wrangell data based on local government result
  mutate(DEMOCRAT = ifelse(county == "Wrangell", 171, DEMOCRAT), 
         REPUBLICAN = ifelse(county == "Wrangell", 526, REPUBLICAN), 
         THIRD = ifelse(county == "Wrangell", 33, THIRD), 
         totalvotes = ifelse(county == "Wrangell", 730, totalvotes)) %>%
  select(-c(state.y, state_po, county_name)) %>%
  rename(state = state.x)
```

The final matched sample dataset incorporates new information as follows: 

  - `totalvotes` is the total number of valid votes in the county
  - `DEMOCRAT` is the number of votes to Democrat candidates in the county
  - `REPUBLICAN` is the number of votes to Republican candidates in the county
  - `THIRD` is the number of votes to third party candidates in the county


## Hispanic or Latino Population

We have efficiently sourced valuable demographic information from the [Annual County Resident Population Estimates by Age, Sex, Race, and Hispanic Origin](https://www.census.gov/programs-surveys/popest/technical-documentation/research/evaluation-estimates/2020-evaluation-estimates/2010s-county-detail.html), meticulously compiled by the U.S. Census Bureau's Vintage 2020 Population Estimates. To streamline our data collection efforts, we opted for a comprehensive nationwide dataset containing generalized county-level data, rather than engaging in individual county-specific collection. 

Focusing on the pivotal years 2020 and 2010, and with a specific interest in the Hispanic or Latino population (regardless of race), we meticulously filtered the dataset. We isolated the data for the years corresponding to July 1, 2020 (identified as `YEAR` == 13) and July 1, 2010 (identified as `YEAR` == 3), targeting the age group encompassing the total population (defined as `AGEGRP` == 0). In adherence to the [official data documentation](https://www2.census.gov/programs-surveys/popest/technical-documentation/file-layouts/2010-2020/cc-est2020-alldata.pdf), we successfully extracted the desired Hispanic population figures, including both males and females. With adept data manipulation, we harmonized these Hispanic variables and integrated them back into our sample dataframe, addressing any discrepancies in county names through a diligent manual correction process.
 
### 2020 Data Collection

```{r}
# Estimates 7/1/2020
hispanic_data_2020 <- 
  read.csv('demo_2010_2020.csv') %>% 
  filter(YEAR == 13, AGEGRP == 0) %>%
  select(STNAME, CTYNAME, H_MALE, H_FEMALE) 

# Remove "County" from names
# Weird character in new mexico county name (not in sample)
hispanic_data_2020 <- hispanic_data_2020[-c(1804),]
hispanic_data_2020$CTYNAME <- sub('County', '', hispanic_data_2020$CTYNAME)
```

```{r}
# Retrieve sample combined info of state and county
sample_st_cty <- (sample %>% mutate(st_cty = paste(state, county, sep="")))$st_cty

# Keep counties in the sample
hispanic_sample_2020 <- hispanic_data_2020 %>%
  mutate(CTYNAME = trimws(CTYNAME, which = "right"),
         n_hispanic = as.numeric(H_MALE) + as.numeric(H_FEMALE),
         st_cty = paste(STNAME, CTYNAME, sep="")) %>%
  filter(st_cty %in% sample_st_cty | CTYNAME %in% c('Hawaii', 'San Francisco', 
                                                    'Wrangell City and Borough', 
                                                    'Emporia city', 'Salem city',
                                                    'Radford city', 'De Witt',
                                                    'Litchfield')) %>%
  # Rename county names that have discrepencies
  mutate(CTYNAME = ifelse(CTYNAME == "Wrangell City and Borough", "Wrangell", CTYNAME), 
         CTYNAME = ifelse(CTYNAME == "De Witt", "DeWitt", CTYNAME),
         CTYNAME = ifelse(CTYNAME == "Emporia city", "Emporia", CTYNAME),
         CTYNAME = ifelse(CTYNAME == "Radford city", "Radford", CTYNAME),
         CTYNAME = ifelse(CTYNAME == "Salem city", "Salem", CTYNAME))

# Join back with sample data
sample_w_hisp2020 <- sample_w_vote %>%
  mutate(state = ifelse(county == "Hawaii", "Hawaii", state)) %>%
  left_join(hispanic_sample_2020, by = join_by("state" == "STNAME", 
                                               "county" == "CTYNAME")) %>%
  select(-c(H_MALE, H_FEMALE, st_cty)) %>%
  rename(hisp_2020 = n_hispanic) %>%
  relocate(hisp_2020, .after = "pop_density")
```

### 2010 Data Collection
    
```{r}
# Estimates 7/1/2010
hispanic_data_2010 <- read.csv('demo_2010_2020.csv') %>% 
  filter(YEAR == 3, AGEGRP == 0) %>%
  select(STNAME, CTYNAME, H_MALE, H_FEMALE) 

# Remove county from names
# Weird ch in new mexico county name, not in sample
hispanic_data_2010 <- hispanic_data_2010[-c(1804),]
hispanic_data_2010$CTYNAME <- sub('County', '', hispanic_data_2010$CTYNAME)
```

```{r}
# Keep counties in the sample
hispanic_sample_2010 <- hispanic_data_2010 %>%
  mutate(CTYNAME = trimws(CTYNAME, which = "right"),
         n_hispanic = as.numeric(H_MALE) + as.numeric(H_FEMALE),
         st_cty = paste(STNAME, CTYNAME, sep="")) %>%
  filter(st_cty %in% sample_st_cty | CTYNAME %in% c('Hawaii', 'San Francisco', 
                                                    'Wrangell City and Borough', 
                                                    'Emporia city', 'Salem city',
                                                    'Radford city', 'De Witt',
                                                    'Litchfield')) %>%
    # Rename county names
  mutate(CTYNAME = ifelse(CTYNAME == "Wrangell City and Borough", "Wrangell", CTYNAME), 
         CTYNAME = ifelse(CTYNAME == "De Witt", "DeWitt", CTYNAME),
         CTYNAME = ifelse(CTYNAME == "Emporia city", "Emporia", CTYNAME),
         CTYNAME = ifelse(CTYNAME == "Radford city", "Radford", CTYNAME),
         CTYNAME = ifelse(CTYNAME == "Salem city", "Salem", CTYNAME))

# Join back with sample data
sample_w_hisp2010 <- sample_w_hisp2020 %>%
  left_join(hispanic_sample_2010, by = join_by("state" == "STNAME", 
                                               "county" == "CTYNAME")) %>%
  select(-c(H_MALE, H_FEMALE, st_cty)) %>%
  rename(hisp_2010 = n_hispanic) %>%
  relocate(hisp_2010, .after = "hisp_2020")
```

The final matched sample dataset incorporates new information as follows: 

  - `hisp_2020` is the population of Hispanic or Latino (any race) in the county in 2020
  - `hisp_2010` is the population of Hispanic or Latino (any race) in the county in 2010


## Exploration: Aging of the U.S. Population 

In the process of framing our exploratory question, we leveraged the rich resources of [official Census data](https://www2.census.gov/programs-surveys/popest/datasets/2010-2020/counties/asrh/CC-EST2020-ALLDATA.csv), which provide detailed demographic information, including age group distributions. The specific question we developed is as follows:

**What is the estimated percentage of the U.S. population aged 65 or above in both 2020 and 2010?**

Our approach involves analyzing the population percentage of seniors with a 10-year interval, a method that promises insights into the evolving age composition in the United States. This analysis sheds light on changes in age demographics and illuminates the challenges and opportunities presented by population aging, which are of paramount importance for future studies and decision-making. To execute this investigation, we have adopted a consistent approach, honing in on the specific age range (`AGEGRP` >= 14) across each county for the selected years. We have further bolstered our analysis by incorporating total population figures by county from 2010, which is instrumental in calculating precise population percentages.
  
### 2010 Population Data by County:

```{r}
pop_2010 <- read.csv('demo_2010_2020.csv') %>% 
  filter(YEAR == 3 & AGEGRP == 0) %>%
  select(STNAME, CTYNAME, TOT_POP)

pop_2010 <- pop_2010[-1804,]
pop_2010$CTYNAME <- sub('County', '', pop_2010$CTYNAME)
```

```{r}
# Keep counties in the sample
pop_sample_2010 <- pop_2010 %>%
  mutate(CTYNAME = trimws(CTYNAME, which = "right"),
         pop.2010 = as.numeric(TOT_POP),
         st_cty = paste(STNAME, CTYNAME, sep="")) %>%
  filter(st_cty %in% sample_st_cty | CTYNAME %in% c('Hawaii', 'San Francisco', 
                                                    'Wrangell City and Borough', 
                                                    'Emporia city', 'Salem city',
                                                    'Radford city', 'De Witt',
                                                    'Litchfield')) %>%
    # Rename county names
  mutate(CTYNAME = ifelse(CTYNAME == "Wrangell City and Borough", "Wrangell", CTYNAME), 
         CTYNAME = ifelse(CTYNAME == "De Witt", "DeWitt", CTYNAME),
         CTYNAME = ifelse(CTYNAME == "Emporia city", "Emporia", CTYNAME),
         CTYNAME = ifelse(CTYNAME == "Radford city", "Radford", CTYNAME),
         CTYNAME = ifelse(CTYNAME == "Salem city", "Salem", CTYNAME))

# Join back
sample_w_pop2010 <- sample_w_hisp2010 %>%
  left_join(pop_sample_2010, by = join_by("state" == "STNAME", 
                                          "county" == "CTYNAME")) %>%
  select(-c(TOT_POP, st_cty)) %>%
  relocate(pop.2010, .after = "pop")
```

### 65+ Age Group Population by County:
    
```{r}
age_65 <- read.csv('demo_2010_2020.csv') %>% 
  filter(YEAR %in% c(3, 13) & AGEGRP >= 14) %>%
  select(STNAME, CTYNAME, YEAR, AGEGRP, TOT_POP)

age_65 <- age_65[-c(18031:18040),]
age_65$CTYNAME <- sub('County', '', age_65$CTYNAME)
```

```{r}
age_sample_65 <- age_65 %>%
  mutate(CTYNAME = trimws(CTYNAME, which = "right"),
         TOT_POP = as.numeric(TOT_POP)) %>%
  group_by(STNAME, CTYNAME, YEAR) %>%
  summarise(pop.65 = sum(TOT_POP), .groups = "drop") %>%
  mutate(YEAR = ifelse(YEAR == 3, "pop_2010_65", "pop_2020_65")) %>%
  pivot_wider(names_from = YEAR, values_from = pop.65) %>%
  mutate(st_cty = paste(STNAME, CTYNAME, sep="")) %>%
  filter(st_cty %in% sample_st_cty | CTYNAME %in% c('Hawaii', 'San Francisco', 
                                                    'Wrangell City and Borough', 
                                                    'Emporia city', 'Salem city',
                                                    'Radford city', 'De Witt',
                                                    'Litchfield')) %>%
  mutate(CTYNAME = ifelse(CTYNAME == "Wrangell City and Borough", "Wrangell", CTYNAME), 
         CTYNAME = ifelse(CTYNAME == "De Witt", "DeWitt", CTYNAME),
         CTYNAME = ifelse(CTYNAME == "Emporia city", "Emporia", CTYNAME),
         CTYNAME = ifelse(CTYNAME == "Radford city", "Radford", CTYNAME),
         CTYNAME = ifelse(CTYNAME == "Salem city", "Salem", CTYNAME))

# Join back
sample_complete <- sample_w_pop2010 %>%
  left_join(age_sample_65, by = join_by("state" == "STNAME", 
                                        "county" == "CTYNAME")) %>%
  select(-c(st_cty, fips)) %>%
  relocate(pop_2020_65, .after = "THIRD") %>%
  rename(pop.2020 = pop)
```

The final matched sample dataset incorporates new information as follows: 

  - `pop.2010` is the total population in each county in 2010
  - `pop_2020_65` is the population of age 65+ in each county in 2020
  - `pop_2010_65` is the population of age 65+ in each county in 2010

# Estimations

The following are the estimations for our questions of interest. We are using the R package [`survey`](https://cran.r-project.org/web/packages/survey/survey.pdf) for the analyses. Since we have some strata with only 1 PSU (lonely PSU), we made these strata certainty PSUs along with D.C. [These PSUs contribute to the estimates themselves, but not to the variance of the estimates](https://r-survey.r-forge.r-project.org/survey/exmample-lonely.html).  

Our estimates consist of totals, means, and ratios. We are going to use a ratio estimator, which estimates the total quantities of both variables then divides, for estimates with less variance. 

```{r}
sample_complete <- 
  read.csv(file = "sample_complete.csv")
# DC certainty PSU
sample_complete[61, 9] = 1
```
  
```{r}
# survey design
options(survey.lonely.psu="certainty")
des <- svydesign(~1, 
                 strata = sample_complete$state, 
                 weights = sample_complete$weights, 
                 fpc = sample_complete$Nh,
                 data = sample_complete)
```

## What is an estimate of the average population density per county in the U.S. in 2020?

```{r}
svymean(~pop_density, des)
confint(svymean(~pop_density, des))
```
We estimate that the average population density per county in the U.S. in 2020 was 180.84 people per square mile. We have a 95% confidence interval of (122.67, 239) for this estimate. 

The actual estimated population density per county is 257.88. This is outside of out 95% confidence interval and therefore indicates some potential errors with our sampling design. In general, our sample does not provide a good estimate for this metric. 

```{r}
mean(clean_county$pop/clean_county$area)
```

## What is an estimate of the total number of people in the U.S. in 2020 who identify as Hispanic or Latino, any race?
  
```{r}
svytotal(~hisp_2020, des)
confint(svytotal(~hisp_2020, des))
```
We estimate that 61,180,804 U.S. residents identified as Hispanic or Latino in 2020. Our 95% confidence interval for this estimate is (40,776,112, 81,585,496). This estimate is very close to the actual estimated value of 61,160,562. 

```{r}
sum(as.numeric(hispanic_data_2020$H_MALE)) + sum(as.numeric(hispanic_data_2020$H_FEMALE))
```

## What is an estimate of the total change in the number of people in the U.S. who identify as Hispanic or Latino, any race, between the 2010 and 2020 censuses?
  
```{r}
# Method 1
sample_change <- sample_complete %>%
  mutate(change = hisp_2020 - hisp_2010)

des1 <- svydesign(~1, 
                 strata = sample_change$state, 
                 weights = sample_change$weights, 
                 fpc = sample_change$Nh,
                 data = sample_change)

svytotal(~change, des1)
confint(svytotal(~change, des1))
```

We estimate that a total of 10,391,602 more U.S. residents identified as Hispanic or Latino in 2020 compared to 2010. We have a 95% confidence interval of (7,249,416, 13,533,788) for this estimate. 

It's clear that a lot more Hispanic/Latino identifying Americans now compared to a decade ago. Out of curiosity, we also wanted to see if the proportion of Hispanic and Latinx residents rose over time. Since the total population data and Hispanic/Latino data came from two different sources, we will use a ratio estimate to estimate both quantities then divide.

```{r}
# Hipanic pop in 2010
svyratio(~hisp_2010, ~pop.2010, des)
confint(svyratio(~hisp_2010, ~pop.2010, des))

# Hipanic pop in 2020
svyratio(~hisp_2020, ~pop.2020, des)
confint(svyratio(~hisp_2020, ~pop.2020, des))
```

We estimate that the percentage of Hispanic and Latino residents in 2010 was 18.17% (95% CI: (13.84%, 22.5%)) while the same statistic in 2020 was estimated to be 20.62% (95% CI: (16.1%, 25.15%)). These results show that we estimate the proportion of Hispanic and Latino residents to be higher in 2020 compared to 2010, but the overlapping confidence intervals suggest this is not a significant difference. 


## What are estimates of the percentages of people in 2020 in the U.S. who voted Republican? How about Democrat? How about a third party?
  
```{r}
svyratio(~REPUBLICAN, ~totalvotes, des)
confint(svyratio(~REPUBLICAN, ~totalvotes, des))
```

We estimate that 48.39% of voters in the 2020 election voted for Trump and the Republican party. We have a 95% confidence interval of (45.02%, 51.77%). Based on the 2020 Presidential Popular Vote Summary, we are 1.54% off from the [actual total of 46.85%](https://www.fec.gov/resources/cms-content/documents/federalelections2020.pdf).

```{r}
svyratio(~DEMOCRAT, ~totalvotes, des)
confint(svyratio(~DEMOCRAT, ~totalvotes, des))
```

We also predict that 49.91% of voters in the 2020 election voted for Biden and the Democratic part. We have a 95% confidence interval of (46.49%, 53.34%). Using the same resource as before, are estimate is off by 1.4%. The [actual percentage is 51.31%](https://www.fec.gov/resources/cms-content/documents/federalelections2020.pdf).

```{r}
svyratio(~THIRD, ~totalvotes, des)
confint(svyratio(~THIRD, ~totalvotes, des))
```

Finally, the expected percentage of voters who voted for a third party in the 2020 election was 1.69%. This estimate had a 95% confidence interval of (1.57%, 1.82%). Given the [actual value of 1.18%](https://www.fec.gov/resources/cms-content/documents/federalelections2020.pdf), we are off by 0.51%. This is a relatively large difference which is reflected by our confidence interval which does not include the true percentage value. Overall, it is clear that our survey design did not adequately estimate this quantity. 

The partisanship results from the 2020 election indicate how close it was. The 95% confidence intervals for the Republican and the Democratic party overlap each other by a great deal. 


## What is the estimated percentage of the U.S. population aged 65 or above in both 2020 and 2010?
  
```{r}
# Proportion of US population age 65+ in 2020
svyratio(~pop_2020_65, ~pop.2020, des)
confint(svyratio(~pop_2020_65, ~pop.2020, des))
```

We estimate that the percentage of U.S. residents who are 65 and over in 2020 was 16.8%. We have a 95% confidence interval of (16.1%, 17.5%) for this estimate. 

```{r}
# Proportion of US population age 65+ in 2010
svyratio(~pop_2010_65, ~pop.2010, des)
confint(svyratio(~pop_2010_65, ~pop.2010, des))
```

We expect that roughly 13.1% of residents in the U.S. were 65 and over in 2010. Our 95% confidence interval for this estimate is (12.6%, 13.7%).  

Given that our confidence intervals do not overlap, we have some evidence to suggest that proportion of residents in the U.S. who are 65 and over was higher in 2020 than in 2010. In other words, people in the U.S. have gotten older overtime (more evidence from [the Census Bureau](https://www.census.gov/newsroom/press-releases/2023/population-estimates-characteristics.html#:~:text=New%20Population%20Estimates%20Highlight%20Increase,by%20the%20U.S.%20Census%20Bureau)). Future studies may seek to model the trajectory of the 65+ age group to determine if this trend is present throughout multiple decades. We may also want to run a regression to see what factors are significantly associated with this pattern while controlling for other variables. These results could aid in understanding the growth of the 65+ age group and what this growth may mean for future health care policies, social security benefits, and so on.  












