---
title: "STA522 Project 1"
author: "Holly Cui"
date: "2023-10-20"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, include=FALSE}
library(tidyverse)
library(tidyr)
library(dplyr)
library(ggplot2)
library(readxl)
library(survey)
library(writexl)
```

```{r data}
county <- read_excel("/Users/hollycui/Desktop/STA\ 522/Project1/county.xlsx", col_names = TRUE)
```

## Study Design

Stratified sampling by 50 states with optimal allocation. 

  - index: row number
  - strata: 50 state* (remove territories + DC as certainty PSU)
  - N: 3246 --> 3144 (after removal)
  - n: 342 --> need explanation*
  - Nh: # of counties in each state
  - nh: # of sampled counties in each state
  
  
## Data Cleaning

```{r}
# remove US territories
US_territories <- c("American Samoa", "Guam", "Northern Mariana Islands", 
                    "Puerto Rico", "U.S. Minor Outlying Islands", 
                    "Virgin Islands (U.S.)")

clean_county <- county %>%
  filter(!state %in% US_territories) %>%
  group_by(state) %>%
  mutate(Nh = n(), Sh = sd(pop)) %>%
  ungroup() %>%
  mutate(id = row_number())
```
  
  
## Sample Size

### Method 1 - Standard

A good maximum sample size is usually around 10% of the population. Thus, we define our sample size as $n = 0.1N = 314$. 

### Method 2 - Fancy but not applicable:

Sample size for known population: 
$$n = \frac{\frac{Z^2 \times p(1-p)}{e^2}}{1+\frac{Z^2 \times p(1-p)}{e^2N}}$$
where: 

  - $Z$: Z-score for the 95% confidence intervals
  - $p$: Standard deviation of how much individual sample data points deviate from the average population
  - $e$: Designated margin of error 
  - $N$: Population size
  
```{r}
# define parameters
Z = 1.96 # for 95% CI
p = 0.5 # default value for expected variance
e = 0.05 # usually around 2% to 5% based on accuracy
N = nrow(clean_county)

# calculate sample size n
n = round((Z^2 * p*(1-p) / e^2) / (1 + (Z^2 * p*(1-p) / (e^2 * N)))) # 342
```


## Optimal Allocation

To calculate the optimal allocation of n=343, we need to adopt the expression as follows:
$$n_h = n\times\frac{\frac{N_hS_h}{N}}{\frac{\Sigma^{H}_{h=1}N_hS_h}{N}}$$
```{r}
# state-level strata statistics (without DC - certainty PSU)
clean_state <- county %>%
  filter(!state %in% US_territories) %>%
  filter(state != "District of Columbia") %>%
  group_by(state) %>%
  summarise(Nh = n(), Sh = sd(pop)) %>%
  ungroup()
```

```{r}
# calculate denominator for optimal allocation
n = 314
denominator = sum(clean_state$Nh * clean_state$Sh)

# summarize nh for each state strata
state_strata <- clean_state %>%
  mutate(nh_round = round((n-1)*Nh*Sh / denominator)) %>%
  mutate(nh = ifelse(nh_round == 0, nh_round+1, nh_round))
         #nh_round = round(nh_true), 
         #diff = nh_round - nh_true) 

# check rounding issue (which doesn't guarantee final n = 341)
# state_strata %>%
#   arrange(desc(diff)) %>%
#   head(sum(state_strata$nh_round) - (n-1))

# fix rounding & get final sampling schema
state_nh <- state_strata %>%
  # mutate(nh_round = ifelse(state == "Tennessee", 6, nh_round), 
  #        nh_round = ifelse(state == "Pennsylvania", 9, nh_round), 
  #        nh_round = ifelse(state == "Illinois", 27, nh_round)) %>%
  select(state, Nh, Sh, nh)
```


## Sampling

```{r}
set.seed(123)

sample_county = c()
for (i in 1:nrow(state_nh)) {
  sample_by_state = sample(clean_county$id[clean_county$state == state_nh$state[i]], 
                           state_nh$nh[i])
  sample_county = c(sample_county, sample_by_state)
}

# add in DC (certainty PSU)
final_sample = c(sample_county, 322)
```

```{r}
sample = clean_county %>%
  filter(id %in% final_sample)
sample

#write.csv(sample, file = "/Users/hollycui/Desktop/STA\ 522/Project1/sample.csv", row.names = FALSE)
```

## Weights & fpc (=Nh)

```{r}
# excluding D.C. for now, its weight is 1
sample_final <- sample %>%
  left_join(state_nh, by = c("state", "Nh", "Sh")) %>%
  select(-id) %>%
  mutate(weights = Nh/nh,
         pop_density = pop/area,
         county = ifelse(grepl("city", county, ignore.case = TRUE),
                         sub(",.*", "", county),
                         county))

#write_xlsx(sample_final, "/Users/hollycui/Desktop/STA\ 522/Project1/sample_final.xlsx")
```


## Collect County-Level Data

  - Matching with FIPS code (change Northwest Hills Planning Region -> Litchfield)
```{r}
# use FIPS for matching
fips <- read.csv("/Users/hollycui/Desktop/STA\ 522/Project1/countyfipstool20190120.csv")
```

```{r}
most_match <- sample_final %>%
  left_join(fips, by = join_by("county" == "cname", "state" == "sname")) %>%
  select(-c(sab, sid, sfips, saint, cfips))

# manually source missing values
# write_xlsx(most_match, "/Users/hollycui/Desktop/STA\ 522/Project1/sample_match.xlsx")
```

```{r}
sample_match <- read_excel("/Users/hollycui/Desktop/STA\ 522/Project1/sample_match.xlsx", col_names = TRUE)
```

  - Voting Data
```{r}
# politic votings dataset
vote <- read.csv("/Users/hollycui/Desktop/STA\ 522/Project1/countypres_2000-2020.csv")
```

```{r}
# create vote dataset by party and county
third_party <- c("OTHER", "GREEN", "LIBERTARIAN")

vote_by_party <- vote %>%
  filter(year == 2020) %>%
  select(state, state_po, county_name, county_fips, candidate, party, candidatevotes, totalvotes) %>%
  mutate(party_group = case_when(
    party %in% third_party ~ "THIRD",
    .default = party
  )) %>%
  select(-c(party, candidate)) %>%
  group_by(state, county_name, party_group) %>%
  summarise(votes = sum(candidatevotes), .groups = "drop")

# clean original data
vote_select <- vote %>%
  filter(year == 2020) %>%
  mutate(party_group = case_when(
    party %in% third_party ~ "THIRD",
    .default = party
  )) %>%
  select(state, state_po, county_name, county_fips, party_group, totalvotes) %>%
  distinct() # remove duplicate rows

# join back
clean_vote <- vote_by_party %>%
  left_join(vote_select, by = join_by(state, county_name, party_group)) %>%
  select(state, state_po, county_name, county_fips, party_group, votes, totalvotes) %>%
  pivot_wider(names_from = party_group, values_from = votes) %>%
  mutate(county_fips = ifelse(state_po == "DC", 11001, county_fips))
```

```{r}
# join final voting results
sample_w_vote <- sample_match %>%
  left_join(clean_vote, by = join_by("fips" == "county_fips")) %>%
  # manual fix Wrangell data based on local government result
  mutate(DEMOCRAT = ifelse(county == "Wrangell", 171, DEMOCRAT), 
         REPUBLICAN = ifelse(county == "Wrangell", 526, REPUBLICAN), 
         THIRD = ifelse(county == "Wrangell", 33, THIRD), 
         totalvotes = ifelse(county == "Wrangell", 730, totalvotes)) %>%
  select(-c(state.y, state_po, county_name)) %>%
  rename(state = state.x)
```

 - Hispanic 2020 Data
 
  - 2020

```{r}
# estimates 7/1/2020
hispanic_data_2020 <- read.csv('/Users/hollycui/Desktop/STA\ 522/Project1/demo_2010_2020.csv') %>% 
  filter(YEAR == 13, AGEGRP == 0) %>%
  select(STNAME, CTYNAME, H_MALE, H_FEMALE) 
```

```{r}
# remove "County" from names
# weird ch in new mexico county name, not in sample
hispanic_data_2020 <- hispanic_data_2020[-c(1804),]
hispanic_data_2020$CTYNAME <- sub('County', '', hispanic_data_2020$CTYNAME)
```

```{r}
sample_st_cty <- (sample %>% mutate(st_cty = paste(state, county, sep="")))$st_cty

# keep counties in the sample
hispanic_sample_2020 <- hispanic_data_2020 %>%
  mutate(CTYNAME = trimws(CTYNAME, which = "right"),
         n_hispanic = as.numeric(H_MALE) + as.numeric(H_FEMALE),
         st_cty = paste(STNAME, CTYNAME, sep="")) %>%
  filter(st_cty %in% sample_st_cty | CTYNAME %in% c('Hawaii', 'San Francisco', 
                                                    'Wrangell City and Borough', 
                                                    'Emporia city', 'Salem city',
                                                    'Radford city', 'De Witt',
                                                    'Litchfield')) %>%
  # rename county names
  mutate(CTYNAME = ifelse(CTYNAME == "Wrangell City and Borough", "Wrangell", CTYNAME), 
         CTYNAME = ifelse(CTYNAME == "De Witt", "DeWitt", CTYNAME),
         CTYNAME = ifelse(CTYNAME == "Emporia city", "Emporia", CTYNAME),
         CTYNAME = ifelse(CTYNAME == "Radford city", "Radford", CTYNAME),
         CTYNAME = ifelse(CTYNAME == "Salem city", "Salem", CTYNAME))
  
```

```{r}
sample_w_hisp2020 <- sample_w_vote %>%
  mutate(state = ifelse(state == "Hawaiʻi", "Hawaii", state)) %>%
  left_join(hispanic_sample_2020, by = join_by("state" == "STNAME", 
                                               "county" == "CTYNAME")) %>%
  select(-c(H_MALE, H_FEMALE, st_cty)) %>%
  rename(hisp_2020 = n_hispanic) %>%
  relocate(hisp_2020, .after = "pop_density")
```


  - 2010
    
```{r}
# estimates 7/1/2010
hispanic_data_2010 <- read.csv('demo_2010_2020.csv') %>% 
  filter(YEAR == 3, AGEGRP == 0) %>%
  select(STNAME, CTYNAME, H_MALE, H_FEMALE) 
  
head(hispanic_data_2010)
```

```{r}
# remove county from names
# weird ch in new mexico county name, not in sample
hispanic_data_2010 <- hispanic_data_2010[-c(1804),]
hispanic_data_2010$CTYNAME <- sub('County', '', hispanic_data_2010$CTYNAME)
```


```{r}
# keep counties in the sample
hispanic_sample_2010 <- hispanic_data_2010 %>%
  mutate(CTYNAME = trimws(CTYNAME, which = "right"),
         n_hispanic = as.numeric(H_MALE) + as.numeric(H_FEMALE),
         st_cty = paste(STNAME, CTYNAME, sep="")) %>%
  filter(st_cty %in% sample_st_cty | CTYNAME %in% c('Hawaii', 'San Francisco', 
                                                    'Wrangell City and Borough', 
                                                    'Emporia city', 'Salem city',
                                                    'Radford city', 'De Witt',
                                                    'Litchfield')) %>%
    # rename county names
  mutate(CTYNAME = ifelse(CTYNAME == "Wrangell City and Borough", "Wrangell", CTYNAME), 
         CTYNAME = ifelse(CTYNAME == "De Witt", "DeWitt", CTYNAME),
         CTYNAME = ifelse(CTYNAME == "Emporia city", "Emporia", CTYNAME),
         CTYNAME = ifelse(CTYNAME == "Radford city", "Radford", CTYNAME),
         CTYNAME = ifelse(CTYNAME == "Salem city", "Salem", CTYNAME))
```


```{r}
sample_w_hisp2010 <- sample_w_hisp2020 %>%
  left_join(hispanic_sample_2010, by = join_by("state" == "STNAME", 
                                               "county" == "CTYNAME")) %>%
  select(-c(H_MALE, H_FEMALE, st_cty)) %>%
  rename(hisp_2010 = n_hispanic) %>%
  relocate(hisp_2010, .after = "hisp_2020")
```


  - Q5 exploration: Aging of population (estimate proportion of age 65+ in 2020 and 2010)
  
    - Get 2010 population data by county:
```{r}
pop_2010 <- read.csv('demo_2010_2020.csv') %>% 
  filter(YEAR == 3 & AGEGRP == 0) %>%
  select(STNAME, CTYNAME, TOT_POP)

head(pop_2010)
```
```{r}
# remove county from names
# weird ch in new mexico county name, not in sample
pop_2010 <- pop_2010[-1804,]
pop_2010$CTYNAME <- sub('County', '', pop_2010$CTYNAME)
```

```{r}
# keep counties in the sample
pop_sample_2010 <- pop_2010 %>%
  mutate(CTYNAME = trimws(CTYNAME, which = "right"),
         pop.2010 = as.numeric(TOT_POP),
         st_cty = paste(STNAME, CTYNAME, sep="")) %>%
  filter(st_cty %in% sample_st_cty | CTYNAME %in% c('Hawaii', 'San Francisco', 
                                                    'Wrangell City and Borough', 
                                                    'Emporia city', 'Salem city',
                                                    'Radford city', 'De Witt',
                                                    'Litchfield')) %>%
    # rename county names
  mutate(CTYNAME = ifelse(CTYNAME == "Wrangell City and Borough", "Wrangell", CTYNAME), 
         CTYNAME = ifelse(CTYNAME == "De Witt", "DeWitt", CTYNAME),
         CTYNAME = ifelse(CTYNAME == "Emporia city", "Emporia", CTYNAME),
         CTYNAME = ifelse(CTYNAME == "Radford city", "Radford", CTYNAME),
         CTYNAME = ifelse(CTYNAME == "Salem city", "Salem", CTYNAME))

# join back
sample_w_pop2010 <- sample_w_hisp2010 %>%
  left_join(pop_sample_2010, by = join_by("state" == "STNAME", 
                                          "county" == "CTYNAME")) %>%
  select(-c(TOT_POP, st_cty)) %>%
  relocate(pop.2010, .after = "pop")
```

    - Get 65+ age group by county:
    
```{r}
age_65 <- read.csv('demo_2010_2020.csv') %>% 
  filter(YEAR %in% c(3, 13) & AGEGRP >= 14) %>%
  select(STNAME, CTYNAME, YEAR, AGEGRP, TOT_POP)
  
head(age_65)
```

```{r}
# remove county from names
# weird ch in new mexico county name, not in sample
age_65 <- age_65[-c(18031:18040),]
age_65$CTYNAME <- sub('County', '', age_65$CTYNAME)
```

```{r}
age_sample_65 <- age_65 %>%
  mutate(CTYNAME = trimws(CTYNAME, which = "right"),
         TOT_POP = as.numeric(TOT_POP)) %>%
  group_by(STNAME, CTYNAME, YEAR) %>%
  summarise(pop.65 = sum(TOT_POP), .groups = "drop") %>%
  mutate(YEAR = ifelse(YEAR == 3, "pop_2010_65", "pop_2020_65")) %>%
  pivot_wider(names_from = YEAR, values_from = pop.65) %>%
  mutate(st_cty = paste(STNAME, CTYNAME, sep="")) %>%
  filter(st_cty %in% sample_st_cty | CTYNAME %in% c('Hawaii', 'San Francisco', 
                                                    'Wrangell City and Borough', 
                                                    'Emporia city', 'Salem city',
                                                    'Radford city', 'De Witt',
                                                    'Litchfield')) %>%
  mutate(CTYNAME = ifelse(CTYNAME == "Wrangell City and Borough", "Wrangell", CTYNAME), 
         CTYNAME = ifelse(CTYNAME == "De Witt", "DeWitt", CTYNAME),
         CTYNAME = ifelse(CTYNAME == "Emporia city", "Emporia", CTYNAME),
         CTYNAME = ifelse(CTYNAME == "Radford city", "Radford", CTYNAME),
         CTYNAME = ifelse(CTYNAME == "Salem city", "Salem", CTYNAME))

# join back
sample_complete <- sample_w_pop2010 %>%
  left_join(age_sample_65, by = join_by("state" == "STNAME", 
                                        "county" == "CTYNAME")) %>%
  select(-c(st_cty, fips)) %>%
  relocate(pop_2020_65, .after = "THIRD") %>%
  rename(pop.2020 = pop)
```


```{r}
#write.csv(sample_complete, file = "/Users/hollycui/Desktop/STA\ 522/Project1/sample_complete.csv", row.names = FALSE)
```













