---
title: "Project 1"
author: "Elyse McFalls and Holly Cui"
date: "2023-10-20"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r library, include=FALSE}
library(tidyverse)
library(tidyr)
library(dplyr)
library(ggplot2)
library(readxl)
library(survey)
```

# Data and Sampling Frame

```{r data}
# loading county data
county <- read_excel("county.xlsx", col_names = TRUE)
```

This project seeks to understand facets the United States using county-level 
data. Specifically, we are interested in examining the population density and 
demographic information Data on the states and counties in the United States 
were collected from the wikipedia page "List of states and territories of the 
United States," which contains the 2020 census estimates for each county's 
population and area 
([source](https://en.wikipedia.org/wiki/List_of_United_States_counties_and_county_equivalents)). 
More detailed data for each county was extracted from the Unites States Census 
Bureau and MIT's Election Data and Science Labs (MEDSL). The data from the 
Census contained demographic estimates for each county from 2010 to 2020
([source](https://www.census.gov/programs-surveys/popest/technical-documentation/research/evaluation-estimates/2020-evaluation-estimates/2010s-county-detail.html)). MEDSL, on the other hand, had data on political participation by county from 2000 to 2020 ([source](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/VOQCHQ)).

All counties from the 50 U.S. states and the District of Columbia were included
in the sampling frame. We did not include counties from U.S. territories because
we are primarily interested in regions that are enfranchised. We also focused in 
data from the years 2010 and 2020 for the analyses. 

```{r}
# removing US territories
US_territories <- c("American Samoa", "Guam", "Northern Mariana Islands", 
                    "Puerto Rico", "U.S. Minor Outlying Islands", 
                    "Virgin Islands (U.S.)")

clean_county <- county %>%
  filter(!state %in% US_territories) %>%
  group_by(state) %>%
  mutate(Nh = n(), Sh = sd(pop)) %>%
  ungroup() %>%
  mutate(id = row_number())
```

# Sampling Procedure

## Sampling Design

In deciding on the sampling design, we wanted to chose a framework that best fit 
the goals of this project. This framework would ideally give us a representative 
sample of the counties in the U.S. that bears in mind the variedness of each 
state. Therefore, we decided on a simple stratified sample with each of the 
50 states serving as individual stratum. Moreover, we are using optimally allocation to 
determine how many counties to sample from each state. The District of Columbia, 
which operates like a city, state, and county, was treated as a state and a county. It was
sampled with 100% certainty ([source](https://statehood.dc.gov/page/faq#:~:text=The%20Constitution%20dictates%20that%20the,4%20million%20people%20and%20growing.)).

By taking a stratified sample, we are ensuring that every state in the U.S. 
is being accounted for. Stratified samples also provide estimates with lower 
standard errors when there is more between variance than within variance. We 
expect this to be the case with counties in each state. For instance, counties 
in Virginia are expected to exhibit more similarities among themselves than when 
compared to counties in Wyoming. However, we also believed that the variance of 
our variables of interest within each state may differ by state. Consider a 
state like North Dakota where most of the counties are rural compared to a state 
like North Carolina that has a mixture of rural and urban areas. We'd anticipate 
that county-level population in North Dakota would be more uniform and the same
statistics in North Carolina would be more varied.
[source](https://mtgis-portal.geo.census.gov/arcgis/apps/storymaps/collections/189aa1dbd64c4c81b3b4a2b71124f6c6?item=1). 
Therefore, instead of sampling proportional to size, we opted to for an optimal 
allocation framework (equation 1). This way, we would sample more from states 
that have high variances for our variables of interest. We decided to focus on 
the populations of each county to optimally allocate the samples of each 
stratum. We expect population size to be related to demographics. Therefore, 
this framework will reduce the variance in our population estimates and likely 
the variance in our demographic related estimates as well.

$$\text{Eq. 1}: n_h = n \frac {N_hS_h/N} {\sum^H_{h=1}N_hS_h/N}$$
$$\text{where N is the total number of counties in the population, n is the sample size,}$$

$$\text{ H is the total number of states, } N_h \text{ is the number of counties in state h, and }$$
$$S_h \text{ is the variance of the county populations in state h}$$

Moreover, taking a simple random sample of counties allows us to get a 
representative sample from each state. We considered taking a sample 
proportional to size, but we felt it was unnecessary to prioritize more 
populated ares. We also decided to sample the District of Columbia with 100% 
certainty because, as the nation's capital, it is an important region and it 
functions like a state and a county. However, if we treated it like a 
state with one county, our optimal allocation formula would assign no samples 
from it due to there being no variance. Therefore, in order to have D.C. in our 
sample, we purposefully included it. 

## Sample Sizes and Weights

We used an initial sampled size of 314, which is 10% of our population. However, 
the output of the optimal allocation formula gave us some states with no counties 
sampled. To have all states accounted for, we rounded any calculated nh values 
less than 0.5 to 1. This left us with a sample of 317. 

The weight for each county is the total number of counties in its respective state 
divided by the number of counties state. For instance, Alabama has 
67 counties and 3 of them were sampled. Therefore, its weight was 22.33. The 
weight for D.C. is 1 since it is a certainty primary sampling unit. 


```{r}
# state-level strata statistics (without DC - certainty PSU)
clean_state <- county %>%
  filter(!state %in% US_territories) %>%
  filter(state != "District of Columbia") %>%
  group_by(state) %>%
  summarise(Nh = n(), Sh = sd(pop)) %>%
  ungroup()
```


```{r}
# calculate denominator for optimal allocation
n = 314
denominator = sum(clean_state$Nh * clean_state$Sh)

# summarize nh for each state strata
state_strata <- clean_state %>%
  mutate(nh_round = round((n-1)*Nh*Sh / denominator)) %>%
  mutate(nh = ifelse(nh_round == 0, nh_round+1, nh_round))
       
# get final sampling schema
state_nh <- state_strata %>%
  select(state, Nh, Sh, nh)
```


```{r}
set.seed(123)

sample_county = c()
for (i in 1:nrow(state_nh)) {
  sample_by_state = sample(clean_county$id[clean_county$state == state_nh$state[i]], 
                           state_nh$nh[i])
  sample_county = c(sample_county, sample_by_state)
}

# add in DC (certainty PSU)
final_sample = c(sample_county, 317)
```

```{r}
sample = clean_county %>%
  filter(id %in% final_sample)
```

```{r}
# excluding D.C. for now, its weight is 1
weights_df <- data.frame(state = state_nh$state, weight = state_nh$Nh/state_nh$nh)
sample <- merge(sample,weights_df, by = "state")
```

The breakdown of the number of counties sampled by state using the optimal 
allocation formula is as follows:

State         | nh           |State         | nh        | State      | nh                     
------------- | -------------| -------------| ----------|----------- | -------
Alabama|3                     | Louisiana|  3             |Ohio| 9
Alaska| 1                     | Maine|  1                 |Oklahoma| 4
Arizona| 8                    | Maryland|  3              |Oregon| 3
Arkansas|2                    | Massachusetts| 4          |Pennsylvania| 9
California|39                | Michigan| 10               |Rhode Island| 1
Colorado| 5                   | Minnesota| 6              |South Carolina|  3     
Connecticut|  1               | Mississippi|  2           |South Dakota | 1                
Delaware| 1                   | Missouri| 7               |Tennessee|6
Florida| 16                    | Montana|  1               |Texas |  48                
Georgia| 11                    | Nebraska| 3               |Utah|3
Hawaii| 1                     | Nevada |  4               |Vermont|1
Idaho| 2                      | New Hampshire|1           |Virginia | 8    
Illinois|  25                  | New Jersey|  3            |Washington|7
Indiana|   5                  | New Mexico| 2             |West Virginia|1
Iowa|  3                      | New York|16                |Wisconsin|5
Kansas|  4                    | North Carolina| 8         |Wyoming|1
Kentucky| 4                | North Dakota|  1          |District of Columbia*| 1

*certainty primary sampling unit


# Results

## Average Population Density per County in the U.S. in 2020

## Amount of Hispanics and Latinos in the U.S. in 2020

## Change in the Amount of Hispanics and Latinos in the U.S. between 2010 and 2020

## Partisan Breakdown in the U.S. in 2020